Data exploration through dot-driven development\footnote{This work was supported by Google Digital News Initiative.}

### 2017-01-05 pivot/paper.tex
Pivot type provider

\usepackage{microtype}%if unwanted, comment out or use option "draft"
\usepackage{pifont}
\usepackage{fp}
\usepackage{graphicx}
\usepackage{semantic}

@@

\newcommand{\lsep}{~|~}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\qident}[1]{\textnormal{\sffamily \guillemotleft #1\guillemotright}}
% \newcommand{\qident}[1]{\textnormal{\sffamily ‹#1›}}

@@

\begin{equation*}
\begin{array}{rclll}
  D & = & \multicolumn{3}{l}{ 
    \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} 
  }\\[0.5em]
  R & = & \Pi_{f_1, \ldots, f_n} (R) &&\textnormal{Projection -- select specified column names}\\
   & \lsep & \sigma_\varphi (R) &&\textnormal{Selection -- filter rows by given predicate}\\
   & \lsep & \tau_{f_1, \ldots, f_n}(R) &&\textnormal{Sorting -- sort by specified columns}\\
   & \lsep & \Phi_{f, \rho_1/f_1, \ldots, \rho_n/f_n} (R) &&\textnormal{Grouping -- group by and calculate aggregates}
   \\[0.5em]
  \rho & = & \ident{count}  &&\textnormal{Count number of rows in the group}\\
   & \lsep & \ident{sum}~f  &&\textnormal{Sum numerical values of the column $f$}\\
   & \lsep & \ident{dist}~f &&\textnormal{Count number of distinct values of the column $f$}\\
   & \lsep & \ident{conc}~f &&\textnormal{Concatenat string values of the column $f$}\\
\\
\end{array} 
\end{equation*}

@@

\begin{figure}
\begin{equation*}
\begin{array}{l}
\Pi_{f_{p(1)}, \ldots, f_{p(m)}} \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} \leadsto \\
  \quad \{f_{p(1)}\mapsto \vect{v_{p(1),1},\ldots,v_{p(1),r}}, \ldots, f_{p(m)}\mapsto \vect{v_{p(m),1},\ldots,v_{p(m),r}}\}  
\\
\\
\sigma_\varphi \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} \leadsto\\
  \quad \{f_1\mapsto \vect{\ldots, v_{1,j}, \ldots}, \ldots, f_n\mapsto \vect{\ldots, v_{n,j},\ldots}\}
  \qquad (\forall j.\;\varphi\, \{f_1\mapsto v_{1,j}, \ldots, f_n\mapsto v_{n,j}\})
\\
\\
\tau_{f_{p(1)}, \ldots, f_{p(m)}} \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} \leadsto\\
  \quad \{f_1\mapsto \vect{v_{1,q(1)},\ldots,v_{1,q(r)}}, \ldots, f_n\mapsto \vect{v_{n,q(1)},\ldots,v_{n,q(r)}}\} 
  \quad \textnormal{where $q$ permutation}\\
  \qquad \textnormal{such that}~i\leq j\implies(v_{p(1),q(i)}, \ldots, v_{p(m),q(i)}) \leq (v_{p(1),q(j)}, \ldots, v_{p(m),q(j)})
  \\
  \\
\Phi_{f_i, \rho_1/f_1', \ldots, \rho_m/f_m'} \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} \leadsto\\
  \quad \{f_1' \mapsto a_1, \ldots, f_m'\mapsto a_m \} 
  \quad \textnormal{where}\\
  \qquad \{g_1,\ldots,g_s\}=\{\{l~|~k\in 1\ldots r, ~v_{i,l}=v_{i,k}\}, ~l\in 1\ldots r\} \\
  ~~\begin{array}{ll}    
    \qquad a_i = \vect{|g_1|, \ldots, |g_s|}  & \textnormal{when}~\rho_i = \ident{count} \\
    \qquad a_i = \vect{\Sigma_{k\in g_1}v_{j,k}, \ldots, \Sigma_{k\in g_s}v_{j,k}} & \textnormal{when}~\rho_i = \ident{sum}~f_j \\
    \qquad a_i = \vect{\Pi_{k\in g_1}v_{j,k}, \ldots, \Sigma_{k\in g_s}v_{j,k}}   & \textnormal{when}~\rho_i = \ident{conc}~f_j \\
    \qquad a_i = \vect{|\{v_{j,k}, k\in g_1\}|, \ldots, |\{v_{j,k}, k\in g_s\}|}  & \textnormal{when}~\rho_i = \ident{dist}~f_j \\
  \end{array}\\
\end{array}
\end{equation*}

@@

\section{Formalising the host language}
\label{sec:foo}

Type providers often provide a thin type-safe layer over a richer untyped runtime component. In case 
of providers for data access (Section~\ref{sec:tps}), the untyped runtime component performs operations
over an external data source. In case of the pivot type provider, the untyped runtime component 
is a relational algebra, which we use for modelling data transformations. We formalize the 
relational algebra in Section~\ref{sec:foo-rel}, together with the object-based host language
in Section~\ref{sec:foo-foo}.

@@

The focus of our work is on data aggregation and so we use a form of relational algebra with 
extensions for grouping and sorting [X,Y,Z]. The syntax is defined in Figure~\ref{fig:foo-rel}.
We write $f$ for column (field) names and we include definition of a data value $D$, which maps
columns to vectors of length $r$ storing the data (values $v$ are defined below).
Aside from standard projection $\Pi$ and selection $\sigma$, our algebra includes sorting 
$\tau$ which takes one or more column names forming the sort key and aggregation $\Phi$, which
requires a single grouping key and several aggregations together with names of the
new columns to be returned.

The semantics of the algbra is given in Figure~\ref{fig:foo-relsem}. We use vector-based semantics
to support sorting and duplicate entries, but otherwise the formalization captures the usual
behaviour. In projection and sorting, we write $f_{p(1)}, \ldots, f_{p(m)}$ to refer to a selection
of fields from $f_1, \ldots, f_n$. Assuming $m\leq n$, $p$ can be seen as a mapping from $\{1\ldots m\}$
to a subset of $\{1\ldots n\}$. In selection, $\varphi$ is a predicate applied to a mapping from 
column names to values. In sorting, we assume that there is a permutation on row indices $q$ such
that the tuples obtained by selecting values according to the sort key are ordered.

The most complex operation is grouping. We need to group data by the value of the column $f_i$ and
then apply aggregations $\rho_1,\ldots,\rho_m$. To do this, we first obtain a set of groups $g_1,\ldots, g_s$
where each group represents a set of indicies of rows belonging to each group. For a given group 
$g_i$ we can then obtain values of column $j$ for rows in the group as $\{v_{j,k}, k\in g_i\}$.
This is used when performing the aggregations -- $|g_i|$ gives the number of rows in the group,
$\Sigma$ sums numerical values and $\Pi$ (a slight notation abuse) concatenates string values.

\subsection{Foo calculus}
\label{sec:foo-foo}

\begin{equation*}
\begin{array}{rcl}
 \tau &=& C \lsep \ident{num} \lsep \ident{string} \lsep \ident{series}\langl \tau_1, \tau_2 \rangl \lsep \ident{Query}\\
 l &=& \kvd{type}~C(x:\tau) = \overline{m} \\[0.0em]
 m &=& \kvd{member}~N:\tau= e
\end{array}
\end{equation*}

\begin{equation*}
\begin{array}{rcl}
 v &\narrow{=}& d \lsep \ident{None} \lsep \ident{Some}(v) \lsep \kvd{new}~C(\overline{v}) \lsep v_1::v_2 \\[0.0em]
 e &\narrow{=}& d \lsep op \lsep e_1~e_2 \lsep \lambda x.e \lsep e.N \lsep \kvd{new}~C(\overline{e}) \\
   &\narrow{|}& \ident{None} \lsep\kvd{match}~e~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\
   &\narrow{|}& \ident{Some}(e) \lsep e_1 = e_2 \lsep {\kvd{if}~e_1~\kvd{then}~e_2~\kvd{else}~e_3} \lsep \kvd{nil} \\
   &\narrow{|}& e_1 :: e_2 \lsep \kvd{match}~e~\kvd{with}~x_1::x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
\end{array}
\end{equation*}

\begin{equation*}
\inference[(foo)]
  {L_1; \Gamma \vdash e : C; L_2 \qquad ((\kvd{type}~C(x:\tau) = ..\;\kvd{member}~N_i : \tau_i = e_i\;..), L) \in L_2}
  {L_1; \Gamma \vdash e.N_i:\tau_i; L_2 \cup L}
\end{equation*}

\newpage

\begin{figure}
\begin{equation*}
\begin{array}{ll}
\ident{main}(F) = C \mapsto (l, L_1 \cup \ldots \cup L_4)\\[0.5em]
\quad l = \kvd{type}~C(x:\ident{Query}) = \\
\qquad \qquad \kvd{member}~\qident{drop columns} : C_1 = C_1(x) &\textnormal{where}~C_1, L_1 = \ident{drop}(F)\\
\qquad \qquad \kvd{member}~\qident{sort data} : C_2 = C_2(x) &\textnormal{where}~C_2, L_2 = \ident{sort}(F)\\
\qquad \qquad \kvd{member}~\qident{group data} : C_3 = C_3(x) &\textnormal{where}~C_3, L_3 = \ident{group}(F)\\
\qquad \qquad \kvd{member}~\qident{get series} : C_4 = C_4(x) &\textnormal{where}~C_4, L_4 = \ident{get-key}(F)\\
\\
\ident{drop}(F) = C, \{l\} \cup \bigcup L_f\\[0.5em]
\quad l = \kvd{type}~C(x:\ident{Query}) = \\
\qquad \qquad \kvd{member}~\ident{then} : C = C(x)                 &\textnormal{where}~C, L = \ident{main}(F)\\
\qquad \qquad \kvd{member}~\qident{drop $f$} : C_f = &\forall f \in \dom(F)~\textnormal{where}~C_f, L_f = \ident{drop}(F')\\
\qquad \qquad \qquad  C_f(\Pi_{\ident{dom}(F')}(x))& \quad\textnormal{and}~ F' = \{ f'\mapsto\tau'\in F, f' \neq f\}\\
\end{array}
\end{equation*}

\caption{Foo}
\label{fig:foo}
\end{figure}

\begin{figure}
\begin{equation*}
\begin{array}{ll}
\ident{get-key}(F) = C, \{l\} \cup \bigcup L_f \\[0.25em]
\quad l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)~\textnormal{where} \\
\qquad \qquad \kvd{member}~\qident{with key $f$} : C_f = C_f(x) &\quad C_f, L_f = \ident{get-val}(F, f)\\
\\
\ident{get-val}(F, f_k) = C, \{l\} \cup \bigcup L_f \\[0.25em]
\quad l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)\setminus\{f\}~\textnormal{where} \\
\qquad \qquad \kvd{member}~\qident{and value $f$} : C_f = C_f(\Pi_{f_k, f_v}(x)) &\quad  C_f~\textnormal{defined as below} \\
\quad l_f = \kvd{type}~C_f(x:\ident{Query}) = &\quad\textnormal{and}~\tau_k, \tau_v~\textnormal{such that} \\
\qquad \qquad \kvd{member}~\ident{series} : \ident{series}\langl\tau_k, \tau_v\rangl = x &\quad \tau_k = F(f_k), \tau_v = F(f) &\\
\end{array}
\end{equation*}

\caption{Foo}
\label{fig:foo}
\end{figure}

@@

Omit paging, because it's boring

Our formalization of the pivot type provider follows the style used by Petricek et al. \cite{fsdata}
in their formalization of F\# Data. In our case, a type provider is a function that takes a schema
$F$ (mapping from field names to field types) and produces a class $C$ together with collection 
of other class defintins $L$ that are used by $C$:
%
\begin{equation*}
\ident{provider}(F) = C, L    ~~\textnormal{where}~F=\{ f_1 \mapsto \tau_1, \ldots, f_n \mapsto \tau_n \}
\end{equation*}
%
The provided class $C$ has a constructor taking \ident{Query}. It may be a class with further 
members that allow refining the query or a class with a single member \ident{series} that returns
a value of type $\ident{series}\langl \tau_1, \tau_2 \rangl$. In the second case, we can evaluate
the provided transformation on input $R$ using an expression $(\kvd{new}~C(R)).\ident{series}$.

\subparagraph{Top-level type.} The top-level type allows the user to select a transformation. The
members return objects with members that allow specifying further properties of each transformation.

\subparagraph{Sorting.} yadda

>>>>

\usepackage{microtype}%if unwanted, comment out or use option "draft"
\usepackage{pifont}
\usepackage{fp}
\usepackage{graphicx}
\usepackage{semantic}
\usepackage{lipsum}

@@

\newcommand{\lsep}{~\,|\,~}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\qident}[1]{\textnormal{\sffamily \guillemotleft #1\guillemotright}}
% \newcommand{\qident}[1]{\textnormal{\sffamily ‹#1›}}

@@

\begin{equation*}
\begin{array}{rclll}
  D & = & \multicolumn{3}{l}{ 
    \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} 
  }\\[0.5em]
  e & = & \Pi_{f_1, \ldots, f_n} (e) &&\textnormal{Projection -- select specified column names}\\
   & \lsep & \sigma_\varphi (e) &&\textnormal{Selection -- filter rows by given predicate}\\
   & \lsep & \tau_{f_1\mapsto\omega_1, \ldots, f_n\mapsto\omega_n}(e) &&\textnormal{Sorting -- sort by specified columns}\\
   & \lsep & \Phi_{f, \rho_1/f_1, \ldots, \rho_n/f_n} (e) &&\textnormal{Grouping -- group by and calculate aggregates}
   \\[0.5em]
  \omega & = & \ident{desc}\lsep\ident{asc}  &&\textnormal{Sort order -- descending or ascending}\\
  \rho & = & \ident{count}  &&\textnormal{Count number of rows in the group}\\
   & \lsep & \ident{sum}~f  &&\textnormal{Sum numerical values of the column $f$}\\
   & \lsep & \ident{dist}~f &&\textnormal{Count number of distinct values of the column $f$}\\
   & \lsep & \ident{conc}~f &&\textnormal{Concatenat string values of the column $f$}\\
\\
\end{array} 
\end{equation*}

@@

\begin{figure}
\begin{equation*}
\begin{array}{l}
\Pi_{f_{p(1)}, \ldots, f_{p(m)}} \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} \leadsto \\
  \quad \{f_{p(1)}\mapsto \vect{v_{p(1),1},\ldots,v_{p(1),r}}, \ldots, f_{p(m)}\mapsto \vect{v_{p(m),1},\ldots,v_{p(m),r}}\}  
\\
\\
\sigma_\varphi \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} \leadsto\\
  \quad \{f_1\mapsto \vect{\ldots, v_{1,j}, \ldots}, \ldots, f_n\mapsto \vect{\ldots, v_{n,j},\ldots}\}
  \qquad (\forall j.\;\varphi\, \{f_1\mapsto v_{1,j}, \ldots, f_n\mapsto v_{n,j}\})
\\
\\
\tau_{f_{p(1)}\mapsto\omega_1, \ldots, f_{p(m)}\mapsto\omega_m} \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} \leadsto\\
  \quad \{f_1\mapsto \vect{v_{1,q(1)},\ldots,v_{1,q(r)}}, \ldots, f_n\mapsto \vect{v_{n,q(1)},\ldots,v_{n,q(r)}}\} 
  \quad \textnormal{where $q$ permutation}\\
  \qquad \textnormal{such that}~\forall i, j.~i\leq j\implies(u_{1,i}, \ldots, v_{m,i}) \leq (v_{1,j}, \ldots, v_{m,j})~ \textnormal{where}\\
  \qquad \quad u_{k,l}=v_{p(k),q(l)} \quad \,~~(\textnormal{when}~\omega_k=\ident{asc}) \\
  \qquad \quad u_{k,l}=-v_{p(k),q(l)} \quad (\textnormal{when}~\omega_k=\ident{desc})
  \\
  \\
\Phi_{f_i, \rho_1/f_1', \ldots, \rho_m/f_m'} \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} \leadsto\\
  \quad \{f_1' \mapsto a_1, \ldots, f_m'\mapsto a_m, f_i\mapsto b \} 
  \quad \textnormal{where}\\
  \qquad \{g_1,\ldots,g_s\}=\{\{l~|~k\in 1\ldots r, ~v_{i,l}=v_{i,k}\}, ~l\in 1\ldots r\} \\
  ~~\begin{array}{ll}    
    \qquad b = \vect{v_{i,k_1},\ldots,v_{i, k_s}} & \textnormal{where}~k_j \in g_j \\
    \qquad a_i = \vect{|g_1|, \ldots, |g_s|}  & \textnormal{when}~\rho_i = \ident{count} \\
    \qquad a_i = \vect{\Sigma_{k\in g_1}v_{j,k}, \ldots, \Sigma_{k\in g_s}v_{j,k}} & \textnormal{when}~\rho_i = \ident{sum}~f_j \\
    \qquad a_i = \vect{\Pi_{k\in g_1}v_{j,k}, \ldots, \Sigma_{k\in g_s}v_{j,k}}   & \textnormal{when}~\rho_i = \ident{conc}~f_j \\
    \qquad a_i = \vect{|\{v_{j,k}\,|\, k\in g_1\}|, \ldots, |\{v_{j,k}\,|\, k\in g_s\}|}  & \textnormal{when}~\rho_i = \ident{dist}~f_j \\
  \end{array}\\
\end{array}
\end{equation*}

@@

\section{Formalising the host language and runtime}
\label{sec:foo}

Type providers often provide a thin type-safe layer over richer untyped runtime components. In case 
of providers for data access (Section~\ref{sec:tps}), the untyped runtime component performs operations
over external data sources. In case of the pivot type provider, the untyped runtime component 
is a relational algebra modelling data transformations. We formalize the relational algebra in 
Section~\ref{sec:foo-rel}, together with the object-based host language in Section~\ref{sec:foo-foo}.

@@

The focus of our work is on data aggregation and so we use a form of relational algebra with 
extensions for grouping and sorting [X,Y,Z]. The syntax is defined in Figure~\ref{fig:foo-rel}.
We write $f$ for column (field) names and we include definition of a data value $D$, which maps
columns to vectors of length $r$ storing the data (values $v$ are defined below).
Aside from standard projection $\Pi$ and selection $\sigma$, our algebra includes sorting 
$\tau$ which takes one or more columns forming the sort key (with sort order $\omega$)
and aggregation $\Phi$, which requires a single grouping key and several aggregations together with 
names of the new columns to be returned.

The semantics of the algbra is given in Figure~\ref{fig:foo-relsem}. We use vector-based semantics
to support sorting and duplicate entries, but otherwise the formalization captures the usual
behaviour. In projection and sorting, we write $f_{p(1)}, \ldots, f_{p(m)}$ to refer to a selection
of fields from $f_1, \ldots, f_n$. Assuming $m\leq n$, $p$ can be seen as a mapping from $\{1\ldots m\}$
to a subset of $\{1\ldots n\}$. In selection, $\varphi$ is a predicate applied to a mapping from 
column names to values. In sorting, we assume that there is a permutation on row indices $q$ such
that the tuples obtained by selecting values according to the given sort key are ordered. The 
auxiliary definition $u_{k,l}$ negates the number to reverse the sort order when descending order
is required.

The most complex operation is grouping. We need to group data by the value of the column $f_i$ and
then apply aggregations $\rho_1,\ldots,\rho_m$. To do this, we first obtain a set of groups $g_1,\ldots, g_s$
where each group represents a set of indicies of rows belonging to each group. For a given group 
$g_i$ we can then obtain values of column $j$ for rows in the group as $\{v_{j,k}\,|\, k\in g_i\}$.
This is used to calculate the resulting data set -- the field $f_i$ becomes a new column formed
by the group keys (obtained by picking one of the indices from $g_j$ for each group); other fields
are calculated by aggregating data in various ways -- $|g_i|$ gives the number of rows in the group,
$\Sigma$ sums numerical values and $\Pi$ (a slight notation abuse) concatenates string values.

\begin{figure}
\begin{equation*}
\begin{array}{rcl}
  v &=& C(v)\,\, \lsep \ident{series}\langl \tau_1, \tau_2 \rangl(v)\; \lsep n \lsep s \lsep D \\
  e &=& C(e)\,\, \lsep \ident{series}\langl \tau_1, \tau_2 \rangl(e)\;\, \lsep x \lsep v \lsep e.N \lsep \ldots\\
  E &=& C(E) \lsep \ident{series}\langl \tau_1, \tau_2 \rangl(E)\, \lsep E.N \\
    &|& \Pi_{f_1, \ldots, f_n}(E) \lsep \sigma_\varphi(E) \lsep \tau_{f_1, \ldots, f_n}(E) \lsep \Phi_{f_i, \rho_1/f_1', \ldots, \rho_m/f_m'} (E) \\
    \\
 \tau &=& C \lsep \kvd{num} \lsep \kvd{string} \lsep \ident{series}\langl \tau_1, \tau_2 \rangl \lsep \ident{Query}\\
 l &=& \kvd{type}~C(x:\tau) = \overline{m} \\
 m &=& \kvd{member}~N:\tau= e
\end{array}
\end{equation*}
\vspace{-1em}

\[
\inference[(\emph{member})~]
{ L(C) = (\kvd{type}~C(x:\tau)= \ldots~\kvd{member}~N_i : \tau_i = e_i~\ldots), L'}
{ (C(v)).N_i \leadsto_L e_i[x \leftarrow v] }\\
\]
\vspace{-2.5em}

\[
\inference[(\emph{context})~]
{e \leadsto_L e'}
{E[e] \leadsto_L E[e']}
\]

\caption{Syntax and remaining reduction rules of the Foo calculus}
\label{fig:foo-syntax}
\end{figure}

\subsection{Foo calculus with lazy context}
\label{sec:foo-foo}

We model the host language using a variant of the Foo calculus [X]. The core of the calculus models
a simple object-based language with objects and members. The syntax of the language is shown in
Figure~\ref{fig:foo-syntax}. The relational algebra defined in Figure~\ref{fig:foo-rel} is included
in the Foo calculus as a model of the runtime components of the pivot type provider -- the values
include the data value $D$ and the expressions include all the operations of the relational algebra.

The Foo calculus includes two special types. \ident{Query} is a type of data and queries 
constructed using the relational algebra. The type $\ident{series}\langl \tau_1, \tau_2 \rangl$ 
models a data series mapping keys of type $\tau_1$ to values of type $\tau_2$. A series is created 
as a result of a query written using the pivot type provider. In the actual implementation, a series 
can be passed to a charting library. A series is a statically-typed wrapper over a \ident{Query} 
value and the proofs in Section~\ref{sec:proofs} show that a series obtained from the pivot type 
provider contains keys and values of matching types.

\subparagraph{Reduction rules.} The reduction relation $\leadsto_L$ is prameterized by a function
$L$ that maps class names to class definitions, together with nested classes associated with the
class definition (used during type-checking as discussed below). The map is not used in the 
reduction rules for the relational algebra, given in Figure~\ref{fig:foo-relsem} and so it was 
omitted there.

The remaining reduction rules are given in Figure~\ref{fig:foo-syntax}. The (\emph{member}) rule 
performs lookup using $L(C)$ to find the definition of the member that is being accessed and then
it reduces member access by substituting the evaluated constructor argument $v$ for a variable $x$.
We assume standard capture-avoiding substitution $[x\leftarrow v]$. The rule ignores the nested 
class definitions $L'$. The (\emph{context}) rule performs reduction in an evaluation context $E$.

\subparagraph{Type checking.} One interesting aspect of type-checking with type providers is that
type providers can provide potentially infinite number of types. In F\#, the types are provided 
lazily as the type checker explores parts of the type space used by the program [X]. Consider:
%
\begin{equation*}
\ident{olympics}.\qident{group data}.\qident{by Athlete}.\qident{sum Gold}.\ident{then}
\end{equation*}

\noindent
The type checker initially knows the type of $\ident{olympics}$ is a class $C_1$ with member 
$\qident{group data}$ and it knows that the type of this member is $C_2$. However, it only needs to 
obtain full definition of $C_2$ when checking the member $\qident{by Athlete}$. Types of other 
member of $C_1$ remain unevaluated when they do not appear in the source code. This aspect of type
providers have been omitted in previous work [X,Y], but it is necessary for the pivot type provider.
The typing rules given are written using the following judgement:
%
\begin{equation*}
L_1; \Gamma \vdash e : \tau; L_2
\end{equation*}

\noindent
The judgement states that given class definitions $L_1$ and a variable context $\Gamma$, the type
of expression $e$ is $\tau$ and the type checking evaluated further class definitions that are
now included in $L_2$. The structure of class definitions $L_1$ is a function mapping a class name 
$C$ to a pair consisting of the definition and a function that provides definitions of 
further classes: 
%
\begin{equation*}
L_1(C) = \kvd{type}~C(x:\tau) = \overline{m}, L'
\end{equation*}

\noindent
The types of members of the class $C$ may be already known classes defined by $L_1$, but also classes 
defined by $L'$. This models laziness as $L'$ is a function that may never be evaluated.

The rules that define type-checking are shown in Figure~\ref{fig:foo-typecheck}. The two rules
that force the discovery of new classes are (\emph{new}) and (\emph{member}). In both cases,
we perform a lookup using $L_2(C)$ which returns a class definition together with further classes
$L$. We treat functions as sets and join $L_2$ with further classes defined by $L$ using $L_2 \cup L$.

\begin{figure}[t]

\[
\inference[(\emph{num})~]{}{L; \Gamma \vdash n : \kvd{num}; L }
\quad
\inference[(\emph{string})~]{}{L; \Gamma \vdash s : \kvd{string}; L}
\quad
\inference[(\emph{var})~]{}{L; \Gamma, x:\tau\vdash x:\tau; L}
\]
\vspace{-2em}

\[
\inference[(\emph{data})~]
  {L; \Gamma \vdash v_{i,j} : \tau;L \quad \tau \in \{ \kvd{num}, \kvd{string} \}}
  {L; \Gamma \vdash \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} : \ident{Query};L }
\]
\vspace{-2em}

\[
\inference[(\emph{proj})~]
  {L_1; \Gamma\vdash e : \ident{Query}; L_2}
  {L_1; \Gamma\vdash \Pi_{f_1, \ldots, f_n} (e) : \ident{Query}; L_2}
\qquad\inference[(\emph{sort})~]
  {L_1; \Gamma\vdash e : \ident{Query}; L_2}
  {L_1; \Gamma\vdash \tau_{f_1, \ldots, f_n}(e): \ident{Query}; L_2}
\]

\[
\inference[(\emph{sel})~]
  {L_1; \Gamma\vdash e : \ident{Query}; L_2}
  {L_1; \Gamma\vdash \sigma_\varphi (e) : \ident{Query}; L_2}
\qquad
\inference[(\emph{group})~]
  {L_1; \Gamma\vdash e : \ident{Query}; L_2}
  {L_1; \Gamma\vdash \Phi_{f, \rho_1/f_1, \ldots, \rho_n/f_n}(e): \ident{Query}; L_2}
\]
\vspace{-2em}

\[
\inference[(\emph{series})~]
  {L_1; \Gamma\vdash e : \ident{Query}; L_2}
  {L_1; \Gamma\vdash \ident{series}\langl \tau_1, \tau_2 \rangl(e) : \ident{series}\langl \tau_1, \tau_2 \rangl; L_2}
\]
\vspace{-2em}

\[
\inference[(\emph{new})~]
  {L_1; \Gamma \vdash e : \tau, L_2 & L_2(C) = (\kvd{type}~C(x:\tau) = \ldots), L}
  {L_1; \Gamma \vdash C(e) : C; L_2 \cup L}
\]
\vspace{-2em}

\[
\inference[(\emph{member})~]
  {L_1; \Gamma \vdash e : C; L_2 \qquad L_2(C)=(\kvd{type}~C(x:\tau) = ..\;\kvd{member}~N_i : \tau_i = e_i\;..), L}
  {L_1; \Gamma \vdash e.N_i:\tau_i; L_2 \cup L}
\]
\vspace{-2em}

\[
\inference[(\emph{class})~]
  { L; \Gamma,x:\tau \vdash e_i : \tau_i; L_{i} & m_i = \kvd{member}~N_i:\tau_i= e_i }
  { L \vdash \kvd{type}~C(x:\tau) = m_1\ldots m_n; L_1 \cup \ldots \cup L_n }
\]
\caption{Type-checking of Foo expressions with lazy context}
\label{fig:foo-typecheck}
\end{figure}

The rules for primitive types and variables are standard.
Input data (\emph{data}) is of type \ident{Query} and all the operations of relational algebra 
take a value of type \ident{Query} and produce a result of type \ident{Query}. This is the 
untyped (or one-typed) subset of the Foo calculus. A \ident{Query} value can be converted into a
series type (\emph{series}) of any type\footnote{This can be seen as a boundary between static and
dynamic typing in gradually typed languages [X,Y,Z].} -- in the pivot type provider, the 
\ident{Query} type is always hidden and users can only work with type-safe series type.

Finally, we also define a judgement $L_1 \vdash l; L_2$ which states that given known classes $L_1$,
a class definition $l$ is well typed and produces known classes $L_2$. This is defined by the
(\emph{class}) rule that checks the types of bodies of all members of the class.

\begin{figure}
\begin{equation*}
\begin{array}{ll}
\ident{pivot}(F) = C, \{ C \mapsto (l, L_1 \cup \ldots \cup L_4) \}\qquad\ball{1}\\[0.5em]
\quad l = \kvd{type}~C(x:\ident{Query}) = \\
\qquad \qquad \kvd{member}~\qident{drop columns} : C_1 = C_1(x) &\textnormal{where}~C_1, L_1 = \ident{drop}(F)\\
\qquad \qquad \kvd{member}~\qident{sort data} : C_2 = C_2(x) &\textnormal{where}~C_2, L_2 = \ident{sort}(F)\\
\qquad \qquad \kvd{member}~\qident{group data} : C_3 = C_3(x) &\textnormal{where}~C_3, L_3 = \ident{group}(F)\\
\qquad \qquad \kvd{member}~\qident{get series} : C_4 = C_4(x) &\textnormal{where}~C_4, L_4 = \ident{get-key}(F)\\
\\
\ident{get-key}(F) = C, \{ C \mapsto (l, \bigcup L_f)\}\qquad\ball{2} \\[0.25em]
\quad l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)~\textnormal{where} \\
\qquad \qquad \kvd{member}~\qident{with key $f$} : C_f = C_f(x) &\quad C_f, L_f = \ident{get-val}(F, f)\\
\\
\ident{get-val}(F, f_k) = \{ C \mapsto (l, \{\}) \}\qquad\ball{3} \\[0.25em]
\quad l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)\setminus\{f_k\}~\textnormal{where} \\
\qquad \qquad \kvd{member}~\qident{and value $f$} : \ident{series}\langl\tau_k, \tau_v\rangl = &\quad  \tau_k = F(f_k),~ \tau_v = F(f) \\
\qquad \qquad \quad \ident{series}\langl\tau_k, \tau_v\rangl(\Pi_{f_k, f}(x))\qquad\ball{4}
\end{array}
\end{equation*}
\vspace{-1em}

\caption{Pivot type provider -- providing entry-point type and accessing transformed data}
\label{fig:tp-main}
\end{figure}

@@

In an implementation, a type provider is an executable component that is called by the compiler
and the editor to provide information about types on demand. In our formalization, we follow the 
style used by Petricek et al. [X], but we add laziness as discussed in Section~\ref{sec:foo-foo}.
We model the core operations (dropping columns, grouping and sorting) in Section~\ref{sec:pivot-core}
and then refine the model to include filtering based on values Section~\ref{sec:pivot-filter}.
We omit paging, as it does not affect the shape of data, but modelling it would require richer 
host language.

\subsection{Pivot type provider}
\label{sec:pivot-core}

A type provider is a function that takes static parameters, such as schema of the input data set, 
and returns a class name $C$ together with a mapping that defines the body of the class, together 
with definitions of associated classes $L$ that may be used by the members of the class $C$. 
In our case, the schema $F$ is a mapping from field names to field types:
%
\begin{equation*}
\begin{array}{l}
\ident{pivot}(F) = C, \{ C \mapsto (\kvd{type}~C(x:\ident{Query}) = \ldots,L) \} \quad \textnormal{where}~F=\{ f_1 \mapsto \tau_1, \ldots, f_n \mapsto \tau_n \}
\end{array}
\end{equation*}
%
The class $C$ provided by the pivot tpe provider has a constructor taking \ident{Query}, which 
represents the, possibly already partly transformed, input data set. It generates members that
allow the user to refine the query and access the data. The type provider is defined using several
helper functions discussed in the rest of this section.

\subparagraph{Entry-point and data access.} Figure~\ref{fig:tp-main} shows three of the functions
defining the pivot type provider. The \ident{pivot} function \ball{1} defines the entry-point type, 
which lets the user choose which operation to perform before specifying parameters of the operation. 
This is the type of \ident{olympics} in the examples throughout this paper. The definition generates 
a new class $C$ with members that wrap the input data in further classes generated by other parts
of the type provider. The result of \ident{pivot} is the class name $C$ together with definition of 
the class and further generated types. The definition is a function that only needs to be evaluated 
when a program accesses a member of the class $C$, modelling the laziness of the type provider.
In the implementation, we return the name $C$, but only need to compute the definition when the 
type checker needs to access the body.
